---
title: "Strong and weak environmental perturbations cause contrasting restructure of ant transportation networks\\linebreak Supplementary Material S2\\linebreak Supplementary Results"
author:
  - Imre Sándor Piross^[sandor.piross@gmail.com]
  - Valentin Lecheval
  - Scott Powell
  - Matina C. Donaldson-Matasci
  - Elva J. H. Robinson
output:
  pdf_document:
    keep_tex: true
    toc: true
bibliography: Supplementary_references.bib
csl: citation_style.csl
header-includes:
    - \usepackage{caption}
---

\captionsetup[table]{labelformat=empty}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(emmeans)
library(knitr)

options(dplyr.summarise.inform = FALSE)
```

```{r data import, cache=TRUE, include=FALSE}
# Directory for simulation experiments
new_dir <- "Simulation run 2023-06-27 15_29_43"

# Gathering simulation statistics
simulation_stats <-
  lapply(list.files(paste0(
    "./Output/", new_dir, "/Network statistics"
  )),
  function(x) {
    readRDS(paste0("./Output/", new_dir, "/Network statistics/",
                   x)) |>
      mutate(dataset = "Simulation",
             colony = x) |>
      relocate(dataset, colony)
  }) |> bind_rows()

# Treatment keys
treatment_keys <-
  data.frame(
    treatment=c("control", "strongest_re", "strongest_sus",
                "weakest_re", "weakest_sus", "random_re", "random_sus"),
    target=factor(c("Control", "Strongest", "Strongest",
             "Weakest", "Weakest","Random","Random"),
             levels = c("Control","Weakest","Random", "Strongest")),
    type=factor(c("Control","Reintroduced","Sustained",
           "Reintroduced","Sustained","Reintroduced","Sustained"),
           levels = c("Control","Sustained","Reintroduced"))
  )

# Adding treatment keys
simulation_stats <- 
simulation_stats |> 
  left_join(treatment_keys, by="treatment") |> 
  relocate(target, type, .after = treatment)

plot_simulation_stat <-
  function(stat, stat_name, t_min=-8) {
    plot_data <- 
      simulation_stats |>
      filter(t>=t_min) |>
      group_by(treatment, type, target, t) |>
      summarise(mean = mean({{stat}}, na.rm = TRUE)) |> 
      filter((!(treatment!="control"&t<(-1))))
    
    # Plot
    plot <-
      plot_data |>
      ggplot(aes(
        y = mean,
        x = t,
        group = treatment,
        colour = target,
        linetype = type
      )) +
      geom_line() +
      ylab(paste("Mean", stat_name,"\n")) +
      scale_colour_manual(values = c("skyblue1", "purple", "red","black"),
                          labels = c("Weakest","Random","Strongest"),
                          breaks = c("Weakest","Random","Strongest"),
                          name="Perturbation target") +
      scale_linetype_manual(values = c("solid", "solid", "dashed"),
                            name="Perturbation type") +
      scale_x_continuous(breaks = seq(from = 0, to = 84, by = 28),
                         minor_breaks = NULL,
                         limits = c(t_min,84)) +
      theme_minimal()+
      theme(
        text=element_text(colour = "black"),
        legend.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom"
        )+
      guides(colour = guide_legend(order = 2), 
             linetype = guide_legend(order = 1))
    
    return(plot)
  }

calculate_means_and_CIs <- 
  function(stat, time_steps=c(0,28,56,84)){
    table <- 
      simulation_stats |>
      filter(t %in% time_steps) |>
      mutate(t=paste0("t=",t)) |> 
      group_by(treatment, type, target, t) |>
      summarise(mean = mean({{stat}}, na.rm = TRUE)) |> 
      pivot_wider(names_from = "t", values_from = "mean") 
    
    # Calculating 95% family-wise CIs
    mod_data <- simulation_stats |>
      filter(t %in% c(84)) |> 
      select({{stat}},treatment) |> 
      rename(STAT = {{stat}})
    
    mod <- lm(STAT~treatment, data = mod_data)
    CIs <-
      confint(ref_grid(mod), adjust = c("BH")) |> 
      select(treatment, lower.CL, upper.CL)
    
    table <- 
    table |> 
      left_join(CIs, by="treatment") |>
      ungroup() |> 
      select(c(-treatment)) |> 
      arrange(target,type)
    
    levels(table$target)[levels(table$target)=="Control"] <- ""
    
    colnames(table)[colnames(table) %in% c("type","target","lower.CL","upper.CL")] <- 
      c("Perturbation type", "Perturbation target",
        "t=84 \\linebreak 95% CI LB", "t=84 \\linebreak 95% CI UB")
    
    
    return(table)
    
  }

# Number of iterations
sim_number <- 
list.files(paste0(
    "./Output/", new_dir, "/Network statistics"
  )) |> length()

```

\newpage
# Overview
This supplementary document provides detailed results for Sandor Simulation MS. The sections present
the results by network measures in the form of a) a figure showing the mean values of the network measures, and b) a table presenting the same means at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons) and 95% family-wise confidence intervals adjusted using the Bonferroni-Holm method @holmSimpleSequentiallyRejective1979
estimating the mean at the end of the simulation ($t = 84$). This interval was included to show the precision reached by running `r sim_number` model simulations. See Supplementary Methods S1.1.4.8.1 Network measures for their detailed description.


\newpage
# S2.1 Number of nests
\renewcommand{\caption}{Figure}
```{r Figure S2.1, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.1: Mean number of nests over time by different treatments."}
plot_simulation_stat(num_nests, "number of nests")
```

```{r Table S2.1}
calculate_means_and_CIs(num_nests) |>
  kable(digits = 3,
        caption = "Table S2.1: Mean number of nests at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.2 Number of used trees
\renewcommand{\caption}{Figure}
```{r Figure S2.2, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.2: Mean number of used trees over time by different treatments."}
plot_simulation_stat(num_trees, "number of used trees")
```

```{r Table S2.2}
calculate_means_and_CIs(num_trees) |>
  kable(digits = 3,
        caption = "Table S2.2: Mean number of used trees at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.3 Number of internest trails
\renewcommand{\caption}{Figure}
```{r Figure S2.3, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.3: Mean number of internest trails over time by different treatments."}
plot_simulation_stat(num_internest, "number of internest trails")
```

```{r Table S2.3}
calculate_means_and_CIs(num_internest) |>
  kable(digits = 3,
        caption = "Table S2.3: Mean number of internest trails at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.4 Number of foraging trails
\renewcommand{\caption}{Figure}
```{r Figure S2.4, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.4: Mean number of foraging trails over time by different treatments."}
plot_simulation_stat(num_foraging, "number of foraging trails")
```

```{r Table S2.4}
calculate_means_and_CIs(num_foraging) |>
  kable(digits = 3,
        caption = "Table S2.4: Mean number of foraging trails at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.5 Trees to nests ratio
\renewcommand{\caption}{Figure}
```{r Figure S2.5, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.5: Mean ratio of the number of trees to the number of nests over time by different treatments."}
plot_simulation_stat(trees_to_nests_rat, "trees to nests ratio")
```

```{r Table S2.5}
calculate_means_and_CIs(trees_to_nests_rat) |>
  kable(digits = 3,
        caption = "Table S2.5: Mean ratio of the number of trees to the number of nests at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.6 Internest trails to nests ratio
\renewcommand{\caption}{Figure}
```{r Figure S2.6, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.6: Mean ratio of the number of internest trails to the number of nests over time by different treatments."}
plot_simulation_stat(internest_to_nests_rat, "internest trails to nests ratio")
```

```{r Table S2.6}
calculate_means_and_CIs(internest_to_nests_rat) |>
  kable(digits = 3,
        caption = "Table S2.6: Mean ratio of the number of internest trails to the number of nests at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.7 Foraging trails to nests ratio
\renewcommand{\caption}{Figure}
```{r Figure S2.7, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.7: Mean ratio of the number of foraging trails to the number of nests over time by different treatments."}
plot_simulation_stat(foraging_to_nests_rat, "foraging trails to nests ratio")
```

```{r Table S2.7}
calculate_means_and_CIs(foraging_to_nests_rat) |>
  kable(digits = 3,
        caption = "Table S2.7: Mean ratio of the number of foraging trails to the number of nests at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.8 Number of network components
\renewcommand{\caption}{Figure}
```{r Figure S2.8, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.8: Mean number of network components over time by different treatments."}
plot_simulation_stat(num_components_nests, "number of network components")
```

```{r Table S2.8}
calculate_means_and_CIs(num_components_nests) |>
  kable(digits = 3,
        caption = "Table S2.8: Mean number of network components at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.9 Network efficiency
\renewcommand{\caption}{Figure}
```{r Figure S2.9, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.9: Mean network efficiency over time by different treatments."}
plot_simulation_stat(network_efficiency, "network efficiency")
```

```{r Table S2.9}
calculate_means_and_CIs(network_efficiency) |>
  kable(digits = 3,
        caption = "Table S2.9: Mean network efficiency at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.10 Network efficiency - nests only
\renewcommand{\caption}{Figure}
```{r Figure S2.10, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.10: Mean efficiency of networks with only nests and internest trails over time by different treatments."}
plot_simulation_stat(network_efficiency_nests, "network efficiency - nests only")
```

```{r Table S2.10}
calculate_means_and_CIs(network_efficiency_nests) |>
  kable(digits = 3,
        caption = "Table S2.10: Mean efficiency of networks with only nests and internest trails at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.11 Network robustness
\renewcommand{\caption}{Figure}
```{r Figure S2.11, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.11: Mean network robustness over time by different treatments."}
plot_simulation_stat(network_robustness, "network robustness")
```

```{r Table S2.11}
calculate_means_and_CIs(network_robustness) |>
  kable(digits = 3,
        caption = "Table S2.11: Mean network robustness at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.12 Network robustness - nests only
\renewcommand{\caption}{Figure}
```{r Figure S2.12, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.12: Mean robustness of networks with only nests and internest trails over time by different treatments."}
plot_simulation_stat(network_robustness_nests, "network robustness - nests only")
```

```{r Table S2.12}
calculate_means_and_CIs(network_robustness_nests) |>
  kable(digits = 3,
        caption = "Table S2.12: Mean robustness of networks with only nests and internest trails at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.13 Network cost
\renewcommand{\caption}{Figure}
```{r Figure S2.13, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.13: Mean network cost over time by different treatments."}
plot_simulation_stat(network_cost, "network cost")
```

```{r Table S2.13}
calculate_means_and_CIs(network_cost) |>
  kable(digits = 3,
        caption = "Table S2.13: Mean network cost at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# S2.14 Network cost - nests only
\renewcommand{\caption}{Figure}
```{r Figure S2.14, fig.width=10, fig.height=7, fig.fullwidth=TRUE, fig.cap="\\medspace S2.14: Mean cost of networks with only nests and internest trails over time by different treatments."}
plot_simulation_stat(network_cost_nests, "network cost - nests only")
```

```{r Table S2.14}
calculate_means_and_CIs(network_cost_nests) |>
  kable(digits = 3,
        caption = "Table S2.14: Mean cost of networks with only nests and internest trails at highlighted timesteps ($t=-8$, the beginning of the burn-in phase; $t=0$, the time of perturbations; $t=28$ the time of tree reintroduction; and $t=56,84$, end of the rest of the seasons). 95% family-wise confidence interval for the mean at the last timestep (t=84).")
```

\newpage
# References
